\secrel{UNIFICATION AS THE SOLUTION OF A SET OF EQUATIONS: A NONDETERMINISTIC
ALGORITHM}

In this section we introduce the basic definitions and give a few theorems which
are useful in proving the correctness of the algorithms. Our ay of stating the
unification problem is slightly more general than the classical one due to
Robinson [18] and directly suggests a number of possible solution methods.

Let
\[A= \bigcup_{i=0,1,..} A_i \quad (A_i \bigcap A_j = \varnothing, i \neq j)\] be
a ranked alphabet, where $A_i$ contains the $i$-adic function symbols
(the elements of $A_0$ are constant symbols). Furthermore, let $V$ be the
alphabet of the variables.
The \term{terms} are defined recursively as follows:

(1) constant symbols and variables are terms;

(2) if $t_1,..,t_n \ (n \geq 1)$ are terms and $f \in A_n$,
then $f(t_1,..,t_n)$ is a term.

A \term{substitution} $\vartheta$ is a mapping from variables to terms,
with $\vartheta(x)=x$ almost everywhere. A substitution can be represented
by a finite set of ordered pairs
$\vartheta={(t_1,x_1),(t_2,x_2),..,(t_m,x_m)}$
where $t_i$ are terms and $x_i$ are distinct variables,
$i = 1,..,m$. To apply a substitution $\vartheta$ to a term $t$, we
simultaneously substitute all occurrences in $t$ of every variable $x_i$ in a
pair $(t_i, x_i)$ of $\vartheta$ with the corresponding term $t_i$. We call the
resulting term $t_\vartheta$.

For instance, given a term $t = f(x_1, g(x_2, a)$ and a substitution 
$\vartheta = {(h(x_2),x_1),(b,x_2)}$, 
we have $t_\vartheta = f(f(x_2),g(b),a)$ and $t_{\vartheta\vartheta} =
f(h(b),g(b),a))$.

The standard unification problem can be written as an equation \[t'=t''\]

A solution of the equation, called a \term{unifier}, is any substitution
$\vartheta$, if it exists, which makes the two terms identical. For instance,
two unifiers of the equation $f(x_1,h(x_1),x_2)=f(g(x_3),x_4,x_3))$ are
$\vartheta_1={(g(x_3),x_1),(x_3,x_2),(h(g(x_3)),x_4)}$ and
$\vartheta_2={(g(a),x_1),(a,x_2),(a,x_3),(h(g(a)),x_4)}$.

In what follows it is convenient also to consider sets of equations
\[t'_j=t''_j, \quad j=1,..,k\]

Again, a \term{unifier} is any substitution which makes all pairs of terms
$t'_j,t''_j$ identical simultaneously.

Now we are interested in finding transformations which produce \emph{equivalent}
sets of equations, namely, transformations which preserve the sets of all
unifiers.
Let us introduce the following two transformations:

\paragraph{(1) Term Reduction.} Let

\begin{equation}\label{mm1}
f(t'_1,t'_2,..,t'_n)=f(t''_1,t''_2,..,t''_n), \quad f \in A_n
\end{equation} 
be an equation where both terms are not variables and where the two root
function symbols are equal. The new set of equations is obtained by replacing
this equation with the following ones: 
\begin{align}\label{mm2}
t'_1 &= t''_1\\
t'_2 &= t''_2\\
&.\\
&.\\
&.\\
t'_n &= t''_n
\end{align}
If $n = 0$, then $f$ is a constant symbol, and the equation is simply erased.

\paragraph{(2) Variable Elimination.} Let $x = t$ be an equation where $x$ is a
variable and $t$ is any term (variable or not). The new set of equations is
obtained by applying the substitution $\vartheta={(t,x)}$ to both terms of all
other equations in the set (without erasing $x = t$).

We can prove the following theorems: 

\paragraph{THEOREM 2.1.} \textit{Let $S$ be a set of equations and let
$f'(t'_1,..t'_n)=f''(t''_1,..,t''_n)$ be an equation of $S$. If $f' \neq f''$,
then $S$ has no unifier. Otherwise, the new set of equations $S'$, obtained by
applying term reduction to the given equation, is equivalent to $S$.}

\paragraph{PROOF.} If $f' \neq f''$, then no substitution can make the
two terms identical.
If $f' = f"$, any substitution which satisfies \ref{mm2} also satisfies
\ref{mm1}, and conversely for the recursive definition of term. $\square$

\paragraph{THEOREM 2.2.} \textit{Let $S$ be a set of equations, and let us apply
variable elimination to some equation $x = t$, getting a new set of equations
$S'$. If variable $x$ occurs in $t$ (but $t$ is not $x$), then $S$ has no
unifier; otherwise, $S$ and $S'$ are equivalent.}

\paragraph{PROOF.} Equation $x = t$ belongs both to $S$ and to $S'$, and thus
any unifier $\vartheta$ (if it exists) of $S$ or of $S'$ must unify $x$ and $t$;
that is, $x_\vartheta$ and $t_\vartheta$ are identical. Now let $t_1 = t_2$ be
any other equation of $S$, and let $t'_1 = t'_2$ be the corresponding equation
in $S'$. Since $t'_1$ and $t'_2$ have been obtained by substituting $t$ for
every occurrence of $x$ in $t_1$ and $t_2$, respectively, we have
$t_{1_\vartheta}=t'_{1_\vartheta}$ and $t_{2_\vartheta}=t'_{2_\vartheta}$. Thus,
any unifier of $S$ is also a unifier of $S'$ and vice versa. Furthermore, if
variable $x$ occurs in $t$ (but $t$ is not $x$), then no substitution
$\vartheta$ can make $x$ and $t$ identical, since $x_\vartheta$ becomes a
subterm of $t_\vartheta$, and thus $S$ has no unifier. $\square$

A set of equations is said to be \term{in solved form} iff it satisfies the
following conditions: 

(1) the equations are $x_j = t_j, j = 1,..,k$;

(2) every variable which is the left member of some equation occurs only there.

A set of equations in solved form has the obvious unifier
\[\vartheta = {(t_1,x_1),(t_2,x_2),..,(t_k,x_k)}\]

If there is any other unifier, it can be obtained as
\[0 = {(t,, x~), (t2, x2) .... , (tk, xk)} U a\]
where a is any substitution which does not rewrite variables xl .... , xk. Thus t~
is called a most general unifier (mgu ).  

The following nondeterministic algorithm shows how a set of equations can be
transformed into an equivalent set of equations in solved form. 

\paragraph{Algorithm 1 }\ \\

Given a set of equations, repeatedly perform any of the following transformations. If no
transformation applies, stop with success. 

(a) Select any equation of the form
\[t=x\]
where t is not a variable and x is a variable, and rewrite it as 
\[x=t.\]

(b) Select any equation of the form
\[X=X\] 
where x is variable, and erase it.  

(c) Select any equation of the form
\[t' = t"\]
where t' and t" are not variables. If the two root function symbols are different, stop with
failure; otherwise, apply term reduction. 

(d) Select any equation of the form
\[x=t\]
where x is a variable which occurs somewhere else in the set of equations and
where t \# x. If x occurs in t, then stop with failure; otherwise, apply variable elimination. 

As an example, let us consider the following set of equations:
\[g(x2) = xl;\]
\[f(xl, h(xl), x2) = f(g(x3), x4, x3).\]

By applying transformation (c) of Algorithm 1 to the second equation we get
\[g(x2) = xl;\]
\[xl = g(x3);\]
\[h(x~) = x4;\]
\[X2 =X3.\]

By applying transformation (d) to the second equation we get
\[g(x2) = g(xs);\]
\[xl = g(x3);\]
\[h(g(x3)) = x4;\]
\[X2 ~- X3.\]

We now apply transformation (c) to the first equation and transformation (a) to
the third equation:
\[X2 ~ X3\]
\[xl = g(x3);\]
\[Xa = h(g(x3));\]
\[X2 ----X3.\]

Finally, by applying transformation (d) to the first equation and transformation
(b) to the last equation, we get the set of equations in solved form:
\[X2 ~- X3 ;\]
\[xl = g(x3);\]
\[x4 = h(g(x3)).\]

Therefore, an mgu of the given system is
\[= {(g(x~), x~), (x3, x2), (h(g(x3)), x4)}.\] 

The following theorem proves the correctness of Algorithm 1. 

\paragraph{THEOREM 2.3}. Given a set of equations S,

(i) Algorithm 1 always terminates, no matter which choices are made.

(ii) If Algorithm 1 terminates with failure, S has no unifier. If Algorithm 1
terminates with success, the set S has been transformed into an equivalent
set in solved form. 
 
\paragraph{PROOF}.\\

(i) Let us define a function F mapping any set of equations S into a triple of
natural numbers (nl, n2, n3). The first number, n~, is the number of variables in
S which do not occur only once as the left-hand side of some equation. The
second number, n2, is the total number of occurrences of function symbols in S.
The third number, n3, is the sum of the numbers of equations in S of type x = x
and of type t = x, where x is a variable and t is not. Let us define a total ordering
on such triples as follows:
\begin{align*}
(n~, n~, n~) > (n~', n2, " n~')\ if\ &n~ > n~'\\
&or\ n~ = n~\ and\ n2 > n2\\
&or\ n~=n" 1 a n- 1 n2 ' ---- n2 "\ and\ n3 ' > n3.\\ 
\end{align*}   

With the above ordering, N 3 becomes a well-founded set, that is, a set where no
infinite decreasing sequence exists. Thus, if we prove that any transformation of
Algorithm 1 transforms a set S in a set S' such that F(S') < F(S), we have
proved the termination. In fact, transformations (a) and (b) always decrease n3
and, possibly, n~. Transformation (c) can possibly increase n3 and decrease nl,
but it surely decreases n2 (by two). Transformation (d) can possibly change n3
and increase n2, but it surely decreases n~. 

(ii) If Algorithm 1 terminates with failure, the thesis immediately follows from
Theorems 2.1 and 2.2. If Algorithm 1 terminates with success, the resulting set of
equations S' is equivalent to the given set S. In fact, transformations (a) and (b)
clearly do not change the set of unifiers, while for transformations (c) and (d) this
fact is stated in Theorems 2.1 and 2.2. Finally, S' is in solved form. In fact, if (a),
(b), and (c) cannot be applied, it means that the equations are all in the form
x = t, with t \# x. If (d) cannot be applied, that means that every v.arialSle
which is the left-hand side of some equation occurs only there.
$\square$\bigskip

The above nondeterministic algorithm provides a widely general version from
which most unification algorithms [2, 3, 7, 13, 15, 16, 18, 22-24] can be derived by
specifying the order in which the equations are selected and by defining suitable
concrete data structures. For instance, Robinson's algorithm [18] might be
obtained by considering the set of equations as a stack. 

