\secrel{Prolog}\label{warren4}\secdown

The language L (resp., the machine M) corresponds to pure Prolog, as it extends
the language L (resp., the machine M) to allow disjunctive definitions.
As in L, an L program is a set of procedure definitions. In L, a definition is
an ordered sequence of clauses (i.e., a sequence of facts or rules) consisting of
all and only those whose head atoms share the same predicate name. That name
is the name of the procedure specified by the definition. L queries are the same
as those of L. The semantics of L operates using top-down leftmost resolution,
an approximation of SLD resolution. Thus, in L, a failure of unification no
longer yields irrevocable abortion of execution but considers alternative choices
of clauses in the order in which they appear in definitions. This is done by chronological
backtracking; i.e., the latest choice at the moment of failure is reexamined
first.

It is necessary to alter M’s design so as to save the state of computation at each
procedure call offering alternatives to restore upon backtracking to this point of
choice. We call such a state a choice point. We thus need to analyze what information
must be saved as a choice point in order to create a record (a choice
point frame) wherefrom a correct state of computation can be restored to offer
another alternative, with all effects of the failed computation undone. Note that
choice point frames must be organized as a stack (just like environments) in order
to reflect the compounding of alternatives as each choice point spawns potentially
more alternatives to try in sequence.

To distinguish the two stacks, let us call the environment stack the AND-stack and
the choice point stack the OR-stack. As with the AND-stack, we organize the
OR-stack as a linked list. The head of this list always corresponds to the latest
choice point, and will be kept in a new global register B, such that upon failure,
computation is made to resume from the state recovered from the choice point
frame indicated by B. When the latest frame offers no more alternatives, it is
popped off the OR-stack by resetting B to its predecessor if one exists, otherwise
computation fails terminally

Clearly, if a definition contains only one clause, there is no need to create a choice
point frame, exactly as was the case in M. For definitions with more than one
alternative, a choice point frame is created by the first alternative; then, it is updated
(as far as which alternative to try next) by intermediate (but non ultimate)
alternatives; finally, it is discarded by the last alternative.

\input{prolog/warren/41env}
\input{prolog/warren/42what}

\secup
